---
title: "Data Tidying"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(readr)
```
package:function eg dylyr::filter
install.packages("dplyr")

cmd + option + i mac
ctrl + alt + i
%>% shortcut Cmd + Shift + M


##Clean data
df_cleaned <- df %>% 
              filter(...) %>%
              select(...)

- select all columns except All and notes since we don't need them
```{r}
catch_original<-read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
head(catch_original)

```
```{r}
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)
```

same as dropping the columns you don't want
```{r}
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
summary(catch_data)
```
-change Chinook to character to numeric
```{r}
catch_clean<-catch_data%>%
  mutate(Chinook=as.numeric(Chinook))
```
###finding spurious characters in Chinook column
which returns row numbers where which is true
```{r}
i<-which(is.na(catch_clean$Chinook))
i
catch_original[i,]
```
```{r}
catch_clean<-catch_data%>%
  mutate(Chinook=if_else(Chinook=="I", "1",Chinook))%>%
  mutate(Chinook=as.numeric(Chinook))
head(catch_clean)
```
##reshaping data
- move from wide to long format
```{r}
catch_long<-catch_clean%>%
  pivot_longer(cols=-c(Region, Year), names_to="Species", values_to="Catch")%>%
  rename(catch_thousands=Catch)%>%  #new name on left, old name on right
  mutate(catch=catch_thousands*1000)%>%
  select(-catch_thousands)
 
head(catch_long)
```

```{r}
catch_wide<-catch_long%>%
  pivot_wider(names_from=Region, values_from=catch)

head(catch_wide)
```

##Split-apply-combine
```{r}
mean_region <- catch_long %>% 
  filter(Species=="Sockeye")%>%
  group_by(Region, Species) %>%
  summarise(catch_mean = mean(catch))

head(mean_region)
```

```{r}
n_region <- catch_long %>% 
  group_by(Region, Species) %>%
  summarize(n = n(),
            .groups="drop")

head(n_region)
```

##joins
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)
```
```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)
nrow(catch_joined)==nrow(catch_long)
```

